Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-15 main.Rmd
cricket and acknowledgement check

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others.

@@

However, even if the situation is not periodic and a similar hierarchy can not be formed, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a  non-temporal data set similar to any temporal data set. There are many interesting questions that can possibly be answered with such a data set irrespective of the type of cyclic granularities. <!--We will explore a few interesting questions and understand how the proposed approach in the paper can help answer some of the questions.-->

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats first or second? Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that players are vulnerable to score through each over and make more runs as they approach towards the end of the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. To answer Q1, winning teams like CSK and MI have indeed different strategies when it comes to batting in the first or second innings.

@@

Another interesting question could be if runs per over set to reduce in subsequent over if fielding is good in the previous over? For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. <!--If a batsman is bowled out, it does not necessarily signify good fielding. So we only include number of catches and run out in an over as a measure of good fielding. 
Difference in runs across overs are likely to be negative if good fielding has an impact on the runs scored in the subsequent overs. --> Dismissals in previous over can lead to a batsman being either defensive or play more vulnerable. Figure \ref{fig:cricex}(b) shows that no dismissals in the previous over will lead to a higher median and quartile spread of run rates compared to the case when there has been at least one dismissals.

@@

Exploratory data analysis is alliteratively finding and summarizing patterns. With temporal data available at more finer scales, exploring periodicity can become overwhelming with so many possible cyclic granularities that could be generated for a context. This work provides a framework to systematically explore distribution of an observed variable across two cyclic time granularities by creating any cyclic granularity, choosing a list of harmonies and thereby recommending possible distribution plots for effective visualization based on relationship and levels of the cyclic granularities.

<!--The infrastructure of systematic exploration of observed variable across pair of cyclic time granularities has been implemented in the
R package `gravitas`. --> A missing piece is to enable user-defined temporal calendars and compute multiple order-up quasi-circular granularities from the hierarchy table. Also, computation of cyclic aperiodic granularities would require computing aperiodic linear granularities first. A few R packages like `almanac` and `gs` provide functionality to create recurring events that are not periodic. These functionalists can be imported in the gravitas package to accommodate for aperiodic cyclic granularities.

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Mon ash University for sharing their wisdom and experience of developing R packages, Dr. Peter Tosca [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). We would also like to thank Nicholas Spyrison for many useful discussions and feedback on the manuscript. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.

>>>>

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

@@

However, even if the situation is not periodic and a similar hierarchy can not be formed, it can be interesting to visualize the distribution of a measured variable across relevant cyclic granularities to throw light on the periodic behavior of a  non-temporal data set similar to any temporal data set. There are many interesting questions that can possibly be answered with such a data set irrespective of the type of cyclic granularities. <!--We will explore a few interesting questions and understand how the proposed approach in the paper can help answer some of the questions.-->

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats first or second? Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that there is no clear upward shift in median and quartile deviation of runs in the second innings as compared to the first innings. This might indicate that players are vulnerable to score through each over and make more runs as they approach towards the end of the first innings, whereas if they bat in the second innings they have a target in mind and are not as vulnerable. Thus winning teams like CSK and MI have indeed different strategies when it comes to batting in the first or second innings.

@@

Another interesting question could be if runs per over set to reduce in subsequent over if fielding is good in the previous over? For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. <!--If a batsman is bowled out, it does not necessarily signify good fielding. So we only include number of catches and run out in an over as a measure of good fielding. 
Difference in runs across overs are likely to be negative if good fielding has an impact on the runs scored in the subsequent overs. --> Dismissals in previous over can lead to a batsman being either defensive or be more resilient. Figure \ref{fig:cricex}(b) shows that no dismissals in the previous over will lead to a higher median and quartile spread of run rates compared to the case when there has been at least one dismissal.

@@

Exploratory data analysis is iteratively finding and summarizing patterns. With temporal data available at more finer scales, exploring periodicity can become overwhelming with so many possible cyclic granularities that could be generated for a context. This work provides a framework to systematically explore distribution of an univariate measured variable across two cyclic time granularities by creating any cyclic granularity, choosing a list of harmonies and thereby recommending possible distribution plots for effective visualization based on relationship and levels of the cyclic granularities.

<!--The infrastructure of systematic exploration of observed variable across pair of cyclic time granularities has been implemented in the
R package `gravitas`. --> A missing piece is to enable user-defined temporal calendars and compute multiple order-up quasi-circular granularities from the hierarchy table. Also, computation of cyclic aperiodic granularities would require computing aperiodic linear granularities first. A few R packages like `almanac` and `gs` provide functionality to create recurring events that are not periodic. These functions can be imported in the `gravitas` package to accommodate for aperiodic cyclic granularities.

@@

The authors would like to thank the cohort at [NUMBATS](https://www.monash.edu/news/articles/team-profile-monash-business-analytics-team), Monash University for sharing their wisdom and experience of developing R packages, Dr. Peter Tosca [from Data61 CSIRO](https://data61.csiro.au/) for providing useful inputs on improving the analysis of smart meter application. The package `gravitas` was built under [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). We would also like to thank Nicholas Spyrison for many useful discussions and feedback on the manuscript. More details about the package 
can be found on the package website https://sayani07.github.io/gravitas/. This article was created with knitr [@knitr, @R-knitr] and R Markdown [@rmarkdown, @R-rmarkdown]. The project's Github repository <https://github.com/Sayani07/paper-gravitas> contains all materials required to reproduce this article and the code is also available online
in the supplemental materials.
